---
output: html_notebook
---

## **Exploratory Data Analysis - Brazilian Imports/Exports**

official data source for brazilian exports e imports, maintened by the government. The dataset contains all trackings of monthly imports and exports of a range of products (soybeans, soybean meal, soybean oil, corn, wheat and sugar), by brazilian states, by routes (air, sea, ground, etc) e from/to which country;

**Importing:**
```{r message=FALSE, warning=FALSE}
# useful library 
# install.packages("tidyverse")

# loading library:
library("tidyverse")

# importing data from csv file:
df <- read.table('data/data_comexstat.csv',
                 header = TRUE,
                 sep = ',',
                 dec = '.')
```

**Examine basic dataset information:**
```{r}
# first rows of dataframe:
head(df)
```

## **Questions:**

##### **Show the evolution of total monthly and total annual exports from Brazil (all states and to everywhere) of ‘soybeans’, ‘soybean oil’ and ‘soybean meal’**

Division in import set and export set:
```{r}
# import set
df_import <- subset(df, df$type == 'Import')
# removing useless column:
df_import <- df_import[-c(5)]
```

```{r}
# export set
df_export <- subset(df, df$type == 'Export')
# removing useless column:
df_export <- df_export[-c(5)]
```

slicing by product
```{r}
# subseting by product
soybeans <- subset(df_export, df_export$product == 'soybeans')
# removing useless column:
soybeans <- soybeans[-c(2, 3 ,4 ,5 ,6)]
# subseting by product:
oil <- subset(df_export, df_export$product == 'soybean_oil')
# removing useless column:
oil <- oil[-c(2, 3 ,4 ,5 ,6)]
# subseting by product
meat <- subset(df_export, df_export$product == 'soybean_meal')
# removing useless column:
meat <- meat[-c(2, 3 ,4 ,5 ,6)]
```

function that creates new columns 
```{r}
YearMonth <- function(df) {
  for (i in 1:nrow(df)) {
    df[i, 'year'] = substr(df[i, 1], 1, 4)
    df[i, 'month'] = substr(df[i, 1], 6, 7)
  }
  return(df)
}

# add new columns to soybeans
soybeans <- YearMonth(soybeans)
soybeans <- soybeans[c("date", "year", "month", "usd")]

# add new columns to oil
oil <- YearMonth(oil)
oil <- oil[c("date", "year", "month", "usd")]

# add new columns to meat
meat <- YearMonth(meat)
meat <- meat[c("date", "year", "month", "usd")]
```

grouping data by year:
```{r message=FALSE, warning=FALSE}
soybeans_year <- soybeans %>% 
  group_by(year) %>% 
    summarize(usd_return = sum(usd))

oil_year <- oil %>% 
  group_by(year) %>% 
    summarize(usd_return = sum(usd))

meat_year <- meat %>% 
  group_by(year) %>% 
    summarize(usd_return = sum(usd))

```

grouping data by month:
```{r message=FALSE, warning=FALSE}
soybeans_month <- soybeans %>% 
  group_by(month) %>% 
    summarize(usd_return = sum(usd))

oil_month <- oil %>% 
  group_by(month) %>% 
    summarize(usd_return = sum(usd))

meat_month <- meat %>% 
  group_by(month) %>% 
    summarize(usd_return = sum(usd))

```

Visualization:
```{r}
# soybeans
ggplot(soybeans_year, aes(x = year, y = usd_return)) +
  geom_col(fill = "lightgreen") +
  labs(title = "Total annual exports from Brazil",
       subtitle = "soybeans",
       x = "Year",
       y = "dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# oil
ggplot(oil_year, aes(x = year, y = usd_return)) +
  geom_col(fill = "green") +
  labs(title = "Total annual exports from Brazil",
       subtitle = "soybean oil",
       x = "Year",
       y = "dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# meat
ggplot(meat_year, aes(x = year, y = usd_return)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Total annual exports from Brazil",
       subtitle = "soybean meat",
       x = "Year",
       y = "dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Visualization 2:
```{r}
# soybeans
ggplot(soybeans_month, aes(x = month, y = usd_return)) +
  geom_col(fill = "lightgreen") +
  labs(title = "Total monthly exports from Brazil",
       subtitle = "soybeans",
       x = "month",
       y = "dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# oil
ggplot(oil_month, aes(x = month, y = usd_return)) +
  geom_col(fill = "green") +
  labs(title = "Total monthly exports from Brazil",
       subtitle = "soybean oil",
       x = "month",
       y = "dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# meat
ggplot(meat_month, aes(x = month, y = usd_return)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Total monthly exports from Brazil",
       subtitle = "soybean meat",
       x = "month",
       y = "dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

##### **What are the 3 most important products exported by Brazil in the last 5 years?**
```{r message=FALSE, warning=FALSE}
# export set
df_export <- subset(df, df$type == 'Export')

# removing useless column:
df_export <- df_export[-c(5)]

# new dataframe of last 5 years:
last5 <- subset(df_export, substr(df$date, 1, 4) == c('2015', '2016', '2017', '2018', '2019'))
# grop by usd return:
product_last5 <- last5 %>% 
  group_by(product) %>% 
    summarize(usd_return = sum(usd))
# sort by dollar return:
product_last5 <- product_last5 %>% arrange(desc(usd_return))
```

Answer visualization:

**Product** | **Usd return** | 
-----:|----:|
**soybeans** | 11537411578| 
**sugar** | 3056523437| 
**soybean_meal** | 1915488713| 


##### **What are the main routes through which Brazil have been exporting ‘corn’ in the last few years? Are there differences in the relative importance of routes depending on the product?**

creating datframe:
```{r}
# creating a new dataframe with cont exportation in the last 5 years:
corn_last5 <- subset(last5, last5$product == 'corn')
# removing useless column:
corn_last5 <- corn_last5[-c(2)]
```

group by route:
```{r}
# counting route use:
best_rote <- corn_last5 %>% 
  count(route)

# sort by dollar return:
best_rote <- best_rote %>% arrange(desc(n))
```
Answer visualization:

**Main rote of corn**|**Number of uses from 2015 to 2019**|
-----:|----:|
**Sea** | 294|

test with another product:
```{r}
last5 %>% 
  group_by(product, route) %>% 
    count(route)
```

Answer visualization 2:

- the most important route of all products is the sea.

#### **Which countries have been the most important trade partners for Brazil in terms of ‘corn’ and ‘sugar’ in the last 3 years?**

```{r message=FALSE, warning=FALSE}
# export set
df_export <- subset(df, df$type == 'Export')

# removing useless column:
df_export <- df_export[-c(5)]

# last 3 years dataset:
last3 <- subset(df_export, substr(df$date, 1, 4) == c('2017', '2018', '2019'))

# filtering by products:
last3_sugar <- subset(last3, last3$product == 'sugar')
last3_corn <- subset(last3, last3$product == 'corn')

# sugar group by usd return:
sugartop <- last3_sugar %>%
  group_by(country) %>%
    summarize(usd_return = sum(usd)) 
  
# corn group by usd return:
corntop <- last3_corn %>%
  group_by(country) %>%
    summarize(usd_return = sum(usd))

# sorting valaues:
sugartop <- sugartop %>% arrange(desc(usd_return))
corntop <- corntop %>% arrange(desc(usd_return))
```
Answer visualization:

**the most important trade partners for Brazil in terms of sugar:**|

**Country** | **Usd return** 
-----:|----:|
China	|235497558			
Algeria	|214996453			
Bangladesh	|160869294	

**the most important trade partners for Brazil in terms of corn:**|

**Country** | **Usd return**
-----:|----:|
Vietnam| 185670299			
Spain| 107624116			
Iran| 98786485		


#### **For each of the products in the dataset, show the 5 most important states in terms of exports:**
```{r message=FALSE, warning=FALSE}
# export set
df_export <- subset(df, df$type == 'Export')

# removing useless column:
df_export <- df_export[-c(5)]

# grouping data by product and country:
important_states <- df_export %>% 
  group_by(product, state) %>% 
    summarize(usd_return = sum(usd))

# slicing by product to plot:
corn_states <- subset(important_states, important_states$product == 'corn')
meal_states <- subset(important_states, important_states$product == 'soybean_meal')
oil_states <- subset(important_states, important_states$product == 'soybean_oil')
soybeans_states <- subset(important_states, important_states$product == 'soybeans')
wheat_states <- subset(important_states, important_states$product == 'wheat')

# sorting:
corn_states <- corn_states %>% arrange(desc(usd_return))
meal_states <- meal_states %>% arrange(desc(usd_return))
oil_states <- oil_states %>% arrange(desc(usd_return))
soybeans_states <- soybeans_states %>% arrange(desc(usd_return))
wheat_states <- wheat_states %>% arrange(desc(usd_return))

# corn plot:
ggplot(corn_states[1:5,], aes(x = state, y = usd_return)) +
  geom_col(fill = "yellow") +
  labs(title = "5 most important states in corn exportation",
       subtitle = "1997-2019",
       x = "State",
       y = "Dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# meal plot:
ggplot(meal_states[1:5,], aes(x = state, y = usd_return)) +
  geom_col(fill = "red") +
  labs(title = "5 most important states in soybean meal exportation",
       subtitle = "1997-2019",
       x = "State",
       y = "Dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# oil plot:
ggplot(oil_states[1:5,], aes(x = state, y = usd_return)) +
  geom_col(fill = "orange") +
  labs(title = "5 most important states in soybean oil exportation",
       subtitle = "1997-2019",
       x = "State",
       y = "Dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# soybean plot:
ggplot(soybeans_states[1:5,], aes(x = state, y = usd_return)) +
  geom_col(fill = "green") +
  labs(title = "5 most important states in soybean exportation",
       subtitle = "1997-2019",
       x = "State",
       y = "Dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# oil plot:
ggplot(wheat_states[1:5,], aes(x = state, y = usd_return)) +
  geom_col(fill = "lightgreen") +
  labs(title = "5 most important states in wheat exportation",
       subtitle = "1997-2019",
       x = "State",
       y = "Dollar")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
**observation:**

- some lines in dataframe have errors like: Cote DIvore,Export,Sea,6000,1
- these errors have been ignored until now because they do not interfere in the general analysis.

#### **What should be the total Brazilian soybeans, soybean_meal, and corn export forecasts, in tons, for the next 11 years (2020-2030) [Yearly]?**

```{r message=FALSE, warning=FALSE}
# export set
df_export <- subset(df, df$type == 'Export')

# removing useless column:
df_export <- df_export[-c(5)]

# add new columns to soybeans
df_forecast <- YearMonth(df_export)

# nwe sort of columns:
df_forecast <- df_forecast[c("date", "year", "month", "product", "state", "country", "route", "tons",  "usd")]

# filter by product
soybeans <- subset(df_forecast, df_forecast$product == "soybeans")
soybean_meal <- subset(df_forecast, df_forecast$product == "soybean_meal")
corn <- subset(df_forecast, df_forecast$product == "corn")

# removing useless column:
soybeans <- soybeans[-c(1, 4)]
soybean_meal <- soybean_meal[-c(1, 4)]
corn <- corn[-c(1, 4)]

# year:
s <- soybeans %>% 
  group_by(year) %>%
    summarize(totalton = sum(tons))

m <- soybean_meal %>% 
  group_by(year) %>%
    summarize(totalton = sum(tons))

c <- corn %>% 
  group_by(year) %>%
    summarize(totalton = sum(tons))

# visualization:
s$year
m$year
c$year

# adding 2003 to m and c dataframes:
df_2003 = data.frame(year=c('2003'), totalton = c(0))

m <- rbind(m[1:6,], df_2003, m[7:22,])
c <- rbind(c[1:6,], df_2003, c[7:22,])
```

in this case I importing coovariates using R studio functions:
```{r}
# basic view
head(covariates)

# filtering by interest range:
covariate_df <- covariates[19:41,]

# new dataframes:
soybeans_df <- covariates[19:41,]
soybeans_df['totalTon'] <- s['totalton']

meal_df <- covariates[19:41,]
meal_df['totalTon'] <- m['totalton']

corn_df <- covariates[19:41,]
corn_df['totalTon'] <- c['totalton']
```
spliting data into train and test:
```{r}
# train:
soybeans_train <- soybeans_df[1:16,]
meal_train <- meal_df[1:16,]
corn_train <- corn_df[1:16,]
# test:
soybeans_test <- soybeans_df[16:23,]
meal_test <- meal_df[16:23,]
corn_test <- corn_df[16:23,]
```
Importing library to use random forest model:
```{r message=FALSE, warning=FALSE}
# install.packages("randomForest")
library(randomForest)

print('I was unable to complete the modeling in the time available, but from here I would use the random forest or some other model to make the predictions')
```


links interessantes:

https://www.r-bloggers.com/how-to-implement-random-forests-in-r/


